---
output:
  html_document:
    keep_md: yes
    number_sections: no
---

```{r chunkoptions, echo=TRUE}
# set global chunk options  

knitr::opts_chunk$set(echo=TRUE)
options(digits=2)

```

# Reproducible Research: Peer Assessment 1
by Melinda Higgins
dated 07/15/2014

## Loading and preprocessing the data

```{r loadData}
# The following code assumes a relative path where 
# the data are located in a subdirectory called "/data/"

activityData <- read.csv("data/activity.csv", header=TRUE)
```

### Summary of the dataset

The data contain `r dim(activityData)[2]` columns of data which are " `r names(activityData)` ".

There are `r dim(activityData)[1]` rows of data covering `r dim(as.data.frame(table(activityData$date)))[1]` days. Each day the number of steps were recorded every 5 minutes during 24 hours for `r 24*60/5` total intervals per day.

### Missing data, Zeros, and Potential Extreme Values (Potential Outliers)

```{r missingReview}
# create a copy of the dataset where the number of steps 
# was NOT missing AND were > 0

nomissing <- activityData[activityData$steps>0 & !is.na(activityData$steps),]
```

* There are `r sum(is.na(activityData$steps))` NA's (missing values);
* There are `r sum(as.vector(activityData$steps==0),na.rm=TRUE)` zeros (0's);
* and `r sum(as.vector(activityData$steps>0),na.rm=TRUE)` entries with non-zero steps ranging from `r min(nomissing$steps)` to `r max(nomissing$steps)` steps in the 5 minute intervals.

Some of these high values seem extreme - possibly too many steps recorded in a 5 minute interval. For example, the maximum number of steps of `r max(nomissing$steps)` in 5 minutes implies making `r max(nomissing$steps)/(5*60)` steps every second.

### Histogram of the Number of Steps Per 5 minute Interval: Including (A) Intervals with Zero steps and Missing Data and (B) For Intervals with the Number of Steps > 0 and no missing data.

```{r hist1, fig.width=6, fig.height=3}
# Create histogram (A) with all the data including 0's and NAs (missing)

hist(activityData$steps, 
     xlab="Number of Steps Per 5 Minute Interval", 
     ylab="Count", 
     main="Histogram of Steps in 5 Minute Intervals")
```

```{r hist2, fig.width=6, fig.height=3}
# Create histogram (B) for only the intervals >0 and no missing

hist(nomissing$steps, 
     xlab="Number of Steps Per 5 Minute Interval", 
     ylab="Count of Steps > 0", 
     main="Histogram of Steps in 5 Minute Intervals (Steps>0)")

```


## What is mean total number of steps taken per day?

Calculate the overall mean and the mean for each day. Also run for the medians per day and overall - with an without the "imputation" method tried below...

```{r summaryTable1}

meanSteps <- tapply(activityData$steps, 
                    activityData$interval, 
                    mean, na.rm=TRUE)
summary(meanSteps)

library(tables)

# this sort of works but won't work with the NAs
tabular((as.factor(interval)+1)~(n=1)+Format(digits=2)*(steps)*(mean + sd), data=nomissing)

# also
aa <- aggregate(activityData$steps, 
                list(activityData$interval), 
                mean, 
                na.rm=TRUE)
names(aa) <- c("Interval","Average Number of Steps")
aa[which.max(aa[,2]),]

# load chron and lattice libraries

library(chron)
library(lattice)

# Use chron() function to create chronological objects for times and days

activitydate <- chron(as.character(activityData$date),
                      format=c(dates="y-m-d"),
                      out.format=c("day months year"))

# create copy fo original dataset, add extracted dates and times
# and then extract the day of the week

activityData2 <- activityData
activityData2$date2 <- activitydate
activityData2$weekday <- weekdays(activityData2$date2)

# now sort over interval and Day of week

aa2 <- aggregate(activityData2$steps, 
                list(activityData2$interval, activityData2$weekday), 
                mean, 
                na.rm=TRUE)

names(aa2) <- c("Interval","Day of Week","Average Number of Steps")
aa2[which.max(aa2[,3]),]

```


## What is the average daily activity pattern?

Look at the raw data - scatterplot by time of day (the increments) - look at overall and by day of the week - also look at a table summarizing these data given the n, mean, median, maybe Q1, Q3, min and max - over all 61 days by time of day and again by day of week. make some plots to go along with...

```{r extractdates, fig.height=12}
# load chron and lattice libraries

#library(chron)
#library(lattice)

# Use chron() function to create chronological objects for times and days

#activitydate <- chron(as.character(activityData$date),
#                      format=c(dates="y-m-d"),
#                      out.format=c("day months year"))

# create copy fo original dataset, add extracted dates and times
# and then extract the day of the week

#activityData2 <- activityData
#activityData2$date2 <- activitydate
#activityData2$weekday <- weekdays(activityData2$date2)

# create an xyplot using the lattice package and ploy Day of Week separately

xyplot(activityData2$steps ~ activityData2$interval | factor(activityData2$weekday),
       type=c("p","spline"),
       col.line="darkorange",
       lwd=3, 
       layout=c(1,7))

```

```{r xtabletest, results='asis'}
# test of xtable function for lm fit summary

library(xtable)
fitlm <- lm(activityData2$steps ~ activityData2$interval)
fitlmxtable <- xtable(fitlm)
print.xtable(fitlmxtable,type="html")

testdataf <- activityData2[1:30,]
testdataf.xtable <- xtable(activityData2[1:30,])
print(testdataf.xtable, type="html", floating=FALSE)


```

### Summary Statistics for Number of Steps in 5 Minute Intervals

```{r xtabletest2, results='asis'}
# second test of xtable using summary statistics

sumsteps <- summary(activityData2$steps)
sumstepsxtable <- xtable(table(sumsteps))
print.xtable(sumstepsxtable, type="html")

```

### Summary Statistics for Number of Steps (in 5 min interval)
=============================================================
Statistic   | Value
----------- | ----------
Mean        | `r round(mean(activityData2$steps, na.rm=TRUE),2)`
Median      | `r round(median(activityData2$steps, na.rm=TRUE),2)`
SD          | `r round(sd(activityData2$steps, na.rm=TRUE),2)`
Min         | `r round(min(activityData2$steps, na.rm=TRUE),2)`
Max         | `r round(max(activityData2$steps, na.rm=TRUE),2)`


## Imputing missing values

Need to consider why the data are "missing" (NA's). These are most likely due to inactivity, although zeros were also recorded during other entries. Perhaps the NAs are due to the monitor being turned off or other reason no data was recorded. 

If we assume that the NA's are due to inactivity then substituting 0's for NA would be appropriate. If this cannot be confirmed, a smoothing approach could be used taking a simple average of the reading before and after the NA is noted. Although if the reading before and after are both NAs then perhaps a zero is best.

However, the number of steps are highly right-skewed with obvious zero-inflation. This distribution indicates that using the mean to substitute for the NAs is not appropriate. Instead using the median would be better - although these are essentially zero for every day as well - again suggesting that substituting zero's for the NAs would be appropriate.

It is noted that without a proper explanation for the source of the NAs, any method used for substitution will be biased. To what extent the chosen method is biased is unverifiable without further information.





## Are there differences in activity patterns between weekdays and weekends?

Need to figure out what function is needed to extract day of the week from the provided dates. Also need to check the date formatting...

